Non-word repetition in Yélî Dnye

### 2020-09-16 manuscript.tex
various bug fixes

We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions. We excluded 3 children who did not have data for one of these two types. As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions. Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
were on average lower than first ones (M = 64, SD = 15), t(38) = 6.30, p = 0. Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.

@@

The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes are even higher M = 100\% (SD = 0\%). The phoneme-based normalized Levenshtein distance was M = 0\% (SD = 0\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted. Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.

@@

We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words. The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.79, SE ß = 0.19, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2. Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).

@@

We next checked whether this may actually be due to frequency of the sounds within the language. We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings ({\textbf{???}}), and applied the natural logarithm. As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses). Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(23)=0.65, =0). Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic frequency (ß = 0.51, SE ß = 0.25, p = 0.04) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.39, SE ß = 0.24, p = 0.11) was not.

@@

\textbf{Option 1: just numbers, first attempts only }
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 0 vowels deleted, and 0 consonants.

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality). As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones, compared to 0 complex ones).

\textbf{Option 2: numbers and proportions, first attempts only}
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets). As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets). Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).

\textbf{Option 3: numbers and proportions, ALL attempts}
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets). As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times about NA\% out of all simple consonant targets). Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).

@@

\begin{tabular}{lll}
\toprule
Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
\midrule
47 (22) & 100 (0) & 0 (0)\\
79 (22) & 100 (0) & 0 (0)\\
78 (19) & 100 (0) & 0 (0)\\
75 (32) & 100 (0) & 0 (0)\\
\bottomrule
\end{tabular}

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals. As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,485.88) = 0.35, p = 0.09). In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.08) = 21.39, p = 0.94), birth order (data missing for 17 children, 2,730.78) = -0.19, p = 0.38) or maternal education (data missing for 2 children, 2,147.84) = 0.07, p = 0.76)

>>>>

We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions. We excluded 3 children who did not have data for one of these two types. As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions. Surprisingly, NWR scores for subsequent repetitions (M = 39, SD = 26)
were on average lower than first ones (M = 65, SD = 15), t(38) = 6.60, p = 0. Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.

@@

The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes are even higher M = 77\% (SD = 10\%). The phoneme-based normalized Levenshtein distance was M = 23\% (SD = 10\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted. Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.

@@

We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words. The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.81, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2. Age was also a significant predictor (ß = 0.32, SE ß = 0.13, p = 0.01).

@@

We next checked whether this may actually be due to frequency of the sounds within the language. We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings ({\textbf{???}}), and applied the natural logarithm. As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses). Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00). Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic frequency (ß = 0.81, SE ß = 0.27, p = 0.00) and age (ß = 0.32, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.00, SE ß = 0.25, p = 1.00) was not.

@@

\textbf{Option 1: just numbers, first attempts only }
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 7 vowels deleted, and 3 consonants.

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels, 31 nasal target vowels produced as oral vowels). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (112 oral vowels produced with a different quality; 13 nasal vowels produced with a different quality). As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 117) than vice versa (2). Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones, compared to 33 complex ones).

\textbf{Option 2: numbers and proportions, first attempts only}
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 7 vowels deleted (about 0.23\% of all vowel targets), and 3 consonants deleted (about 0.12\% of all consonant targets).

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels or about 0.98\% out of all oral vowel targets, 31 nasal target vowels produced as oral vowels or about 2.37\% out of all nasal vowel targets). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (112 oral vowels produced with a different quality or about 1.82\% out of all oral vowel targets; 13 nasal vowels produced with a different quality or about 2.04\% out of all nasal vowel targets). As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 117 times or about 1.95\% out of all complex consonant targets) than vice versa (2 times or about 1.95\% out of all simple consonant targets). Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones or about 0.38\% out of all simple consonant targets, compared to 33 complex ones or about 0.99\% out of all complex consonant targets).

\textbf{Option 3: numbers and proportions, ALL attempts}
Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 12 vowels deleted (about 0.28\% of all vowel targets), and 6 consonants deleted (about 0.19\% of all consonant targets).

As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 1.10\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 2.53\% out of all nasal vowel targets). Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (205 oral vowels produced with a different quality or about 2.16\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 2.36\% out of all nasal vowel targets). As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or about 2.43\% out of all complex consonant targets) than vice versa (2 times about 2.43\% out of all simple consonant targets). Simple consonants were mispronounced as other simple consonants quite frequently (123 simple consonants mispronounced as other simple ones or about 0.53\% out of all simple consonant targets, compared to 55 complex ones or about 0.98\% out of all complex consonant targets).

@@

\begin{tabular}{lll}
\toprule
Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
\midrule
47 (22) & 59 (17) & 41 (17)\\
79 (22) & 92 (9) & 8 (9)\\
78 (19) & 93 (7) & 7 (7)\\
74 (32) & 91 (12) & 9 (12)\\
\bottomrule
\end{tabular}

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals. As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,483.87) = 0.35, p = 0.09). In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.13) = 21.37, p = 0.90), birth order (data missing for 17 children, 2,600.01) = -0.13, p = 0.54) or maternal education (data missing for 2 children, 2,045.68) = 0.11, p = 0.61)
